\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage[margin=1in]{geometry}
\usepackage[]{amsthm} 
\usepackage[]{amsmath} 
\usepackage[]{amssymb} 

\title{Multicast file sharing -- Report}
\author{Kevin Wu, Sam Zhang}
\date{Due: April 6, 2026}

\begin{document}
\maketitle

\setlength{\parindent}{0ex} 
\setlength{\parskip}{5pt} % Adjusts the space between paragraphs

\section*{Overview}


We introduce a \texttt{NACK}-oriented reliable multicast file-sharing protocol that supports full file transfer for both existing receivers and late joiners with low network and host overhead. The sender is effectively stateless and cyclically transmits control packets with metadata for all files. Receivers are stateful: they track received chunks and request missing chunk ranges through \texttt{NACK} packets. Integrity is checked at two levels: per-chunk during reception and file-level at completion. The protocol gives high reliability and fast catch-up for new joiners, but worst-case receiver swarms can still generate substantial repair traffic.

\section*{Protocol Design}

Our protocol uses three UDP multicast packet types: \texttt{METADATA}, \texttt{DATA}, and \texttt{NACK}. All packets share a small header with packet type and payload size. \texttt{METADATA} carries file id, file size, chunk count, chunk size, file checksum, and filename. \texttt{DATA} carries one chunk plus its chunk checksum. \texttt{NACK} carries an inclusive missing-chunk range for repair. The overall model is sender-driven multicast with receiver-driven selective repair.


\subsection*{Sender}

Sender has two core responsibilities:

\begin{enumerate}
	\item It cyclically sends metadata packets for all files so both existing receivers and late joiners can synchronize file state.

	\item It keeps a receive channel open for \texttt{NACK} packets. When a \texttt{NACK} arrives, it validates the file id and retransmits only the requested missing sequence range.
\end{enumerate}

This keeps idle load low (only periodic metadata cycles, e.g., every 3 seconds) while still supporting new or reconnected joiners. Once a joiner receives metadata, it requests exactly what it is missing via \texttt{NACK}, causing a temporary burst of retransmission only when recovery is needed.

\subsection*{Receiver}

Receivers populate per-file trackers on \texttt{METADATA}. Trackers keep progress state, including received-chunk count and missing chunks. On metadata reception events, receivers run a repair check and send \texttt{NACK}s for missing ranges when needed.

Receivers also use backoff to reduce \texttt{NACK} implosion:

1. At startup, each tracker waits briefly before issuing NACKs.

2. Before sending \texttt{NACK}s, receivers back off and check whether another receiver already requested the same file, then suppress or delay their own requests.

This keeps repair traffic selective while controlling synchronized NACK bursts.

Worst-case load happens when receivers join one-by-one and each new receiver joins right after the previous receiver is fully served. Then retransmissions are not shared, so each receiver effectively triggers another full-file repair window. Let $S$ be file size (bytes) and $R$ be the number of such late/reconnected receivers:

$$
L_{\text{worst}} \approx (R+1)S
$$

where $S$ is the original multicast and $RS$ is worst-case repair traffic. Ignoring small control overhead, total load scales linearly with staggered joiners.

\subsection*{Comparison to FLUTE}

A different multicast protocol is FLUTE (File Delivery over Unidirectional Transport), where the sender has no receive channel. It continuously cycles all data chunks, and reliability comes from eventual consistency, as missing chunks are recovered in later cycles.

We like FLUTE's simplicity, but we did not choose it because its load is generally higher when little retransmission is needed, for example with no new joiners or when receivers miss only a few chunks. Our main requirement is minimizing unnecessary control and data traffic, so we use a \texttt{NACK}-driven selective-repair protocol.

\section*{Integrity Verification}

\subsection*{Chunk-Level Checksum}

Chunk-level checksum is computed when each \texttt{DATA} packet is generated and carried with that packet. The receiver validates every incoming chunk before accepting it. If chunk-level checksum fails, the receiver treats it as bad data and immediately re-requests that chunk range via \texttt{NACK}.

\subsection*{File-Level Combined Checksum}

File-level checksum is computed at sender initialization (before transmission) and included in metadata. After a receiver has all chunks for a file, it computes a combined file checksum locally and compares it to the expected metadata value.
The combination is done incrementally over chunks: each chunk checksum is mixed with its sequence number and payload length (including a sequence-dependent rotate), then XORed into a running checksum.

This is a two-step guard for file integrity: chunk-level during reception, then file-level at completion. If file-level verification fails, we assume chunks were valid on receipt and corruption happened locally during or after assembly. Since it is hard to pinpoint the corrupted local chunk(s), we \texttt{NACK} the whole file range and re-fetch the full file.




\section*{Performance Benchmark}

\section*{Limitations}

\end{document}
